---
title: "Getting Started with causalGBKMR"
author: "[Authors]"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with causalGBKMR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  eval = requireNamespace("causalGBKMR", quietly = TRUE)
)
```

## Overview

**causalGBKMR** implements the g-BKMR method for estimating causal effects of
time-varying environmental mixtures. It combines:

- **g-computation** to handle time-varying confounding
- **Bayesian Kernel Machine Regression (BKMR)** to flexibly model nonlinear and
  non-additive exposure-response relationships

This vignette walks through the full workflow in three scenarios:

1. **Basic analysis** --- small data, standard BKMR engine
2. **Custom interventions** --- user-defined exposure contrasts
3. **Large-sample analysis** --- using fastBKMR for computational efficiency

---

## 1. Installation

### Step 1: Install R dependencies from CRAN

```{r install-cran, eval=FALSE}
install.packages(c("bkmr", "dplyr", "fields"))
```

### Step 2: Install causalGBKMR from GitHub

```{r install-pkg, eval=FALSE}
# If you don't have devtools or remotes:
install.packages("devtools")

# Install causalGBKMR
devtools::install_github("Lyric98/causalGBKMR")
```

### Step 3 (optional): Install fastBKMR for large datasets

Only needed if your sample size is large (n > 2000). See Scenario 3 below.

```{r install-fbkmr, eval=FALSE}
remotes::install_github("junwei-lu/fbkmr")
```

### Verify installation

```{r load-packages, message=FALSE, warning=FALSE}
library(causalGBKMR)
```

---

## 2. Background: What Problem Does g-BKMR Solve?

In environmental health studies we often have:

- **Multiple correlated exposures** (e.g., arsenic, lead, cadmium) measured at
  **multiple time points**
- **Time-dependent confounders** (e.g., BMI, blood pressure) that are affected
  by past exposures and also influence the outcome
- Interest in **the causal effect of the entire exposure mixture history** on a
  health outcome

Standard regression would suffer from time-varying confounding bias.
The g-formula addresses this, and BKMR provides the flexible
exposure-response surface. g-BKMR puts them together.

**The causal estimand:**

$$
\text{ATE} = E[Y^{\bar{a}^*}] - E[Y^{\bar{a}}]
$$

where $\bar{a}$ and $\bar{a}^*$ are two exposure histories (by default, all
exposures at the 25th vs. 75th percentile).

---

## 3. Preparing Your Data

### What you need

causalGBKMR expects three inputs:

| Input | Type | Dimensions | Description |
|-------|------|-----------|-------------|
| **Y** | numeric vector | length $n$ | Outcome (continuous or binary) |
| **Z** | numeric matrix | $n \times (p \times T)$ | Exposures: $p$ mixture components at $T$ time points |
| **X** | numeric matrix | $n \times (L \times T + B)$ | $L$ time-dependent covariates at $T$ time points, plus $B$ baseline covariates |

### How to organize your columns

**Exposure matrix Z** --- group by time point, then by component:

```
[Mix1_t0, Mix2_t0, ..., Mixp_t0, Mix1_t1, ..., Mixp_t(T-1)]
```

**Covariate matrix X** --- time-dependent covariates first, then baseline:

```
[TDcov1_t1, TDcov2_t1, ..., TDcovL_t1, ..., TDcovL_t(T-1), Baseline1, ..., BaselineB]
```

> **Important:** Exposures (Z) include baseline $t=0$ through $t=T-1$.
> Time-dependent covariates (X) start at $t=1$ --- they are measured *after*
> baseline exposure. Baseline covariates (e.g., sex, age) go at the end of X.

### Data preparation tips

Before running causalGBKMR, we recommend the following checks:

```{r data-tips, eval=FALSE}
# 1. Check for missing data -- g-BKMR can handle some NAs but complete
#    cases are preferred
cat("Missing in Y:", sum(is.na(Y)), "\n")
cat("Missing in Z:", sum(is.na(Z)), "\n")
cat("Missing in X:", sum(is.na(X)), "\n")

# 2. Check exposure distributions -- environmental data is often right-skewed.
#    The package log-transforms by default (log_transform_mixtures = TRUE).
#    If your data is already log-transformed, set log_transform_mixtures = FALSE.
apply(Z, 2, summary)

# 3. Check for zero or negative values in exposures -- log(0) is undefined.
#    The package handles this by adding a small constant, but it's better to
#    address it beforehand (e.g., replace with LOD/sqrt(2)).
cat("Zeros in Z:", sum(Z == 0, na.rm = TRUE), "\n")
cat("Negatives in Z:", sum(Z < 0, na.rm = TRUE), "\n")

# 4. Check correlations between exposures -- high correlation is expected
#    (and handled by BKMR), but extreme collinearity (r > 0.99) may cause
#    numerical issues.
cor_mat <- cor(Z, use = "complete.obs")
cat("Max pairwise correlation:", max(abs(cor_mat[upper.tri(cor_mat)])), "\n")

# 5. Check sample size relative to model complexity.
#    Rule of thumb: you need at least 10 observations per predictor in the
#    largest BKMR model (the outcome model, which has p*T + L*(T-1) Z columns).
n_predictors_outcome <- ncol(Z) + ncol(X) - n_baseline
cat("Outcome model has", n_predictors_outcome, "predictors\n")
cat("Suggested minimum n:", 10 * n_predictors_outcome, "\n")
```

### Converting from common data formats

If your data is in long format (one row per person-visit), you need to reshape
to wide format first:

```{r long-to-wide, eval=FALSE}
library(tidyr)

# Example: long-format data with columns: id, visit, arsenic, lead, bmi, sex, outcome
wide_data <- long_data %>%
  pivot_wider(
    id_cols = c(id, sex, outcome),
    names_from = visit,
    values_from = c(arsenic, lead, bmi),
    names_sep = "_t"
  )

# Then extract matrices
Z <- as.matrix(wide_data[, c("arsenic_t0", "lead_t0", "arsenic_t1", "lead_t1")])
X <- cbind(
  as.matrix(wide_data[, c("bmi_t1")]),   # TD covariates (start at t1)
  wide_data$sex                            # baseline
)
Y <- wide_data$outcome
```

---

## 4. Scenario 1: Basic Analysis (Standard BKMR)

This is the standard use case for moderate sample sizes ($n < 2000$).

### Step 1: Generate example data

```{r scenario1-data}
set.seed(123)
n <- 200
T_points <- 3   # 3 time points (t=0, t=1, t=2)
p <- 2           # 2 exposures (e.g., arsenic, lead)
Ldim <- 1        # 1 time-dependent confounder (e.g., waist circumference)
n_baseline <- 1  # 1 baseline covariate (e.g., sex)

# Outcome
Y <- rnorm(n, mean = 50, sd = 10)

# Exposure matrix: 2 metals x 3 time points = 6 columns
# Using lognormal to mimic environmental concentrations
Z <- matrix(rlnorm(n * p * T_points, meanlog = 1, sdlog = 0.5),
            nrow = n, ncol = p * T_points)

# Covariate matrix: 1 TD covariate x 3 time points + 1 baseline = 4 columns
waist_base <- rnorm(n, 85, 12)
X <- cbind(
  waist_base + rnorm(n, 1, 2),    # waist at t=1
  waist_base + rnorm(n, 2, 2),    # waist at t=2
  waist_base + rnorm(n, 3, 2),    # waist at t=3
  rbinom(n, 1, 0.5)               # sex (baseline)
)

cat("Y:", length(Y), "observations\n")
cat("Z:", nrow(Z), "x", ncol(Z), "(", p, "exposures x", T_points, "time points)\n")
cat("X:", nrow(X), "x", ncol(X), "(", Ldim, "TD cov x", T_points, "times +",
    n_baseline, "baseline)\n")
```

### Step 2: Prepare data with `prepare_gbkmr_data()`

This function converts your matrices into the internal data frame format. It
handles variable naming, optional log-transformation, and input validation.

```{r scenario1-prepare}
prepared_data <- prepare_gbkmr_data(
  Y = Y,
  Z = Z,
  X = X,
  time_points = T_points,
  mixture_components = p,
  td_covariates = Ldim,
  baseline_covariates = n_baseline,
  td_covariate_names = "waist",        # give meaningful names to TD covariates
  log_transform_mixtures = TRUE         # log-transform (recommended for raw concentrations)
)

# Inspect the result
head(prepared_data[, 1:8])
```

The output data frame has columns in this order: baseline covariates, baseline
TD covariates, exposures at each time point, TD covariates at each time point,
outcome, and subject id.

### Step 3: Verify variable detection

Before running the analysis, check that the package correctly detected your
variable structure:

```{r scenario1-detect}
detection <- detect_variable_patterns(prepared_data, T = T_points)
print(detection)
```

This should show:

- `p` = number of exposures per time point
- `Ldim` = number of time-dependent covariates per time point
- The base names of your TD covariates

If the detection is wrong, check your column naming or provide explicit
parameters to `run_gbkmr_panel()` instead of using `gbkmr_run()`.

### Step 4: Run g-BKMR analysis

**For real analyses**, use production-quality settings:

```{r scenario1-run-production, eval=FALSE}
results <- gbkmr_run(
  data = prepared_data,
  outcome = "Y",
  outcome_type = "continuous",
  time_points = T_points,
  iter = 15000,                        # MCMC iterations (at least 10000)
  n = min(500, nrow(prepared_data)),   # subsample for BKMR fitting
  K = 1000,                            # Monte Carlo samples for g-computation
  n_knots = 50,                        # kernel approximation knots
  engine = "bkmr",                     # standard BKMR
  verbose = TRUE
)

print(results)
summary(results)
```

For this vignette, we use small parameters so it finishes in seconds:

```{r scenario1-run-tiny}
results <- gbkmr_run(
  data = prepared_data,
  outcome = "Y",
  outcome_type = "continuous",
  time_points = T_points,
  iter = 500,       # tiny for demo only!
  n = 50,
  K = 10,
  n_knots = 15,
  engine = "bkmr",
  verbose = TRUE
)
```

### Step 5: Interpret results

```{r scenario1-results}
# The causal effect estimate (ATE)
cat("Average Treatment Effect:", round(results$causal_effect$estimate, 4), "\n")
cat("95% Credible Interval: (",
    round(results$causal_effect$lower, 4), ",",
    round(results$causal_effect$upper, 4), ")\n")

# Counterfactual means
cat("\nE[Y | low exposure]:", round(results$counterfactual_means$low, 4), "\n")
cat("E[Y | high exposure]:", round(results$counterfactual_means$high, 4), "\n")

# Variable importance (beta coefficients from outcome model)
cat("\nVariable importance (outcome model betas):\n")
print(round(results$variable_importance, 4))
```

### Step 6: Diagnostic plots

Always check convergence before trusting results.

```{r scenario1-diagnostics, fig.width=7, fig.height=4}
raw <- results$raw_results

# Posterior distribution of ATE
diff_posterior <- raw$Yastar - raw$Ya
hist(diff_posterior, breaks = 30, col = "steelblue", border = "white",
     main = "Posterior Distribution of ATE",
     xlab = "ATE (high - low exposure)")
abline(v = results$causal_effect$estimate, col = "red", lwd = 2)
abline(v = 0, col = "gray40", lty = 2, lwd = 2)
legend("topright", c("Estimate", "Zero"), col = c("red", "gray40"),
       lty = c(1, 2), lwd = 2, bty = "n")
```

```{r scenario1-trace, fig.width=7, fig.height=3.5}
# Trace plots for convergence
par(mfrow = c(1, 2))
plot(raw$Ya, type = "l", col = "steelblue",
     main = "E[Y | low exposure]", ylab = "Y(a)", xlab = "MCMC iteration")
plot(raw$Yastar, type = "l", col = "tomato",
     main = "E[Y | high exposure]", ylab = "Y(a*)", xlab = "MCMC iteration")
par(mfrow = c(1, 1))
```

**What to look for:** Trace plots should look like "hairy caterpillars" ---
stationary with no trends. If you see drift or slow mixing, increase `iter`.

---

## 5. Scenario 2: Custom Intervention Levels

By default, g-BKMR compares all exposures at the 25th percentile (low) vs.
75th percentile (high). You can customize this in two ways.

### Option A: Change the quantile probabilities

```{r scenario2-quantiles, eval=FALSE}
# Compare 10th vs 90th percentile (wider contrast)
results_wide <- gbkmr_run(
  data = prepared_data,
  outcome = "Y",
  outcome_type = "continuous",
  time_points = T_points,
  a_probs = c(0.10, 0.90),      # 10th vs 90th percentile
  iter = 15000, n = 500, K = 1000
)
```

### Option B: Specify exact exposure values

Use this when you want to compare specific exposure scenarios (e.g., a
regulatory standard vs. observed levels). You need `run_gbkmr_panel()` for this:

```{r scenario2-custom, eval=FALSE}
# Build named vectors matching exposure column names
exposure_names <- paste0("logM", rep(1:2, 3), "_", rep(0:2, each = 2))

# Low scenario: all exposures at log(1) (i.e., concentration = 1)
a_low <- setNames(rep(0, length(exposure_names)), exposure_names)

# High scenario: all exposures at log(5) (i.e., concentration ~ 5)
a_high <- setNames(rep(log(5), length(exposure_names)), exposure_names)

results_custom <- run_gbkmr_panel(
  sim_popn = prepared_data,
  T = T_points,
  p = 2,
  mediator_basenames = "waist",
  common_covariates = c("sex", "waist_0"),
  a_vals = a_low,
  astar_vals = a_high,
  iter = 15000, n = 500, K = 1000, n_knots = 50
)

cat("ATE:", results_custom$diff_gBKMR, "\n")
```

---

## 6. Scenario 3: Large-Sample Analysis with fastBKMR

When sample size is large ($n > 2000$), standard BKMR becomes computationally
expensive because the Gaussian process kernel scales as $O(n^3)$. **fastBKMR**
(via the `fbkmr` package) addresses this by splitting data into subsets, fitting
BKMR on each in parallel, and combining results.

### When to use fastBKMR

| Sample size | Recommendation |
|-------------|---------------|
| $n < 1000$ | Standard BKMR (`engine = "bkmr"`) |
| $1000 < n < 3000$ | Either works; try BKMR with `n_knots` first |
| $n > 3000$ | Use fastBKMR (`engine = "fastbkmr"`) |

### Install fastBKMR

```{r scenario3-install, eval=FALSE}
# fastBKMR is not on CRAN -- install from GitHub
remotes::install_github("junwei-lu/fbkmr")

# Verify
library(fbkmr)
?skmbayes
```

### Running with fastBKMR

The interface is **identical** to standard BKMR --- just change `engine`:

```{r scenario3-run, eval=FALSE}
# Suppose you have a large prepared dataset (n = 6000)
results_fast <- gbkmr_run(
  data = large_prepared_data,
  outcome = "Y",
  outcome_type = "continuous",
  time_points = 3,
  engine = "fastbkmr",           # <-- the only change needed
  n_subset = 10,                 # split data into 10 subsets (~600 obs each)
  n_cores = 10,                  # parallel cores (ideally one per subset)
  iter = 15000,
  n = nrow(large_prepared_data), # use all data (fastBKMR handles the scaling)
  K = 1000,
  verbose = TRUE
)

print(results_fast)
```

### Key parameters for fastBKMR

| Parameter | Default | Description |
|-----------|---------|-------------|
| `engine` | `"bkmr"` | Set to `"fastbkmr"` to enable |
| `n_subset` | `10` | Number of data subsets. More subsets = faster per-subset fitting, but slightly more variance |
| `n_cores` | `10` | Parallel cores. Set to `n_subset` for maximum parallelism |

**Tip on choosing `n_subset`:** Each subset should have at least ~200
observations. So for $n = 6000$, `n_subset = 10` gives ~600 per subset
(good). For $n = 2000$, try `n_subset = 5` to keep ~400 per subset.

### How it works under the hood

1. `fbkmr::skmbayes()` splits the data into `n_subset` random subsets
2. Standard BKMR (`bkmr::kmbayes()`) is fit independently on each subset, in
   parallel across `n_cores` cores
3. During g-computation, the package calls `bkmr::SamplePred()` on a
   **randomly selected subset** at each MCMC iteration
4. Results (betas, predictions) are aggregated across subsets

> **Technical note:** `bkmr::SamplePred()` works on individual subset fits
> (class `bkmrfit`) but fails on the full `skmbayes` list object. causalGBKMR
> handles this automatically --- no user action needed.

### Advanced: using `run_gbkmr_panel()` directly

For full control over model fitting, you can bypass `gbkmr_run()` and call the
core function:

```{r scenario3-panel, eval=FALSE}
results_fast <- run_gbkmr_panel(
  sim_popn = large_prepared_data,
  T = 3,
  p = 4,
  mediator_basenames = c("waist", "bmi"),
  common_covariates = c("sex", "waist_0", "bmi_0"),
  engine = "fastbkmr",
  n_subset = 10,
  n_cores = 10,
  n = 6000,
  iter = 15000,
  sel = seq(12000, 15000, by = 25),   # post-burn-in samples
  K = 1000,
  verbose_every = 20
)
```

---

## 7. Tips and Best Practices

### Parameter tuning guide

| Goal | Adjustment |
|------|-----------|
| **Faster prototyping** | Reduce `iter` (5000), `K` (200), `n` (200) |
| **More stable estimates** | Increase `K` (2000) and `length(sel)` |
| **Memory issues** | Reduce `n_knots` (30) or `n` |
| **Large dataset** | Switch to `engine = "fastbkmr"` |
| **Binary outcome** | Set `outcome_type = "binary"` |

### Recommended production settings

```{r production, eval=FALSE}
# Standard BKMR (n < 2000)
results <- gbkmr_run(
  data = prepared_data,
  time_points = T_points,
  iter = 20000,
  n = min(800, nrow(prepared_data)),
  K = 1000,
  n_knots = 50,
  engine = "bkmr"
)

# fastBKMR (n > 2000)
results <- gbkmr_run(
  data = prepared_data,
  time_points = T_points,
  iter = 15000,
  n = nrow(prepared_data),   # use full data
  K = 1000,
  engine = "fastbkmr",
  n_subset = 10,
  n_cores = 10
)
```

### Common pitfalls

1. **Forgetting to log-transform exposures.** Environmental concentrations are
   typically right-skewed. `prepare_gbkmr_data()` sets
   `log_transform_mixtures = TRUE` by default. If your data is *already*
   log-transformed, set it to `FALSE` to avoid double-transforming.

2. **Too few MCMC iterations.** With `iter = 5000`, BKMR may not converge.
   Use at least `iter = 15000` for real analyses. Always check trace plots
   (see Scenario 1, Step 6).

3. **`sel` exceeds `iter`.** If you set `sel = seq(22000, 24000, ...)` but
   `iter = 15000`, you will get an error. The package auto-calculates `sel`
   when using `gbkmr_run()`, but with `run_gbkmr_panel()` you must ensure
   `max(sel) <= iter`.

4. **Knot warnings.** Messages like "Number of nearest neighbors reduced" from
   `fields::cover.design()` are harmless when $n$ is small relative to
   `n_knots`. Reduce `n_knots` if they bother you.

5. **Columns below the limit of detection (LOD).** For values below LOD,
   replace with `LOD / sqrt(2)` before passing to causalGBKMR. Do not leave
   as zero (log(0) is undefined).

---

## 8. Function Reference

| Function | Purpose | When to use |
|----------|---------|-------------|
| `prepare_gbkmr_data()` | Convert matrices (Y, Z, X) to g-BKMR format | Always --- first step |
| `validate_user_matrices()` | Check input dimensions and types | Debugging data issues |
| `detect_variable_patterns()` | Auto-detect variable structure | Before running, to verify setup |
| `gbkmr_run()` | High-level analysis wrapper | Recommended for most users |
| `run_gbkmr_panel()` | Core g-BKMR with full parameter control | Advanced: custom interventions, fine-tuning |

---

## Session Info

```{r session-info}
sessionInfo()
```
