---
title: "Getting Started with causalGBKMR"
author: "[Authors]"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with causalGBKMR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  eval = requireNamespace("causalGBKMR", quietly = TRUE)
)
```

## Overview

**causalGBKMR** implements the g-BKMR method for estimating causal effects of
time-varying environmental mixtures. It combines:

- **g-computation** to handle time-varying confounding
- **Bayesian Kernel Machine Regression (BKMR)** to flexibly model nonlinear and
  non-additive exposure-response relationships

This vignette walks through the full workflow in three scenarios:

1. **Basic analysis** --- small data, standard BKMR engine
2. **Custom interventions** --- user-defined exposure contrasts
3. **Large-sample analysis** --- using fastBKMR for computational efficiency

---

## 1. Installation

### Step 1: Install R dependencies from CRAN

```{r install-cran, eval=FALSE}
install.packages(c("bkmr", "dplyr", "fields"))
```

### Step 2: Install causalGBKMR from GitHub

```{r install-pkg, eval=FALSE}
# If you don't have devtools or remotes:
install.packages("devtools")

# Install causalGBKMR
devtools::install_github("Lyric98/causalGBKMR")
```

### Step 3 (optional): Install fastBKMR for large datasets

Only needed if your sample size is large (n > 2000). See Scenario 3 below.

```{r install-fbkmr, eval=FALSE}
remotes::install_github("junwei-lu/fbkmr")
```

### Verify installation

```{r load-packages, message=FALSE, warning=FALSE}
library(causalGBKMR)
```

---

## 2. Background: What Problem Does g-BKMR Solve?

In environmental health studies we often have:

- **Multiple correlated exposures** (e.g., arsenic, lead, cadmium) measured at
  **multiple time points**
- **Time-dependent confounders** (e.g., BMI, blood pressure) that are affected
  by past exposures and also influence the outcome
- Interest in **the causal effect of the entire exposure mixture history** on a
  health outcome

Standard regression would suffer from time-varying confounding bias.
The g-formula addresses this, and BKMR provides the flexible
exposure-response surface. g-BKMR puts them together.

**The causal estimand:**

$$
\text{ATE} = E[Y^{\bar{a}^*}] - E[Y^{\bar{a}}]
$$

where $\bar{a}$ and $\bar{a}^*$ are two exposure histories (by default, all
exposures at the 25th vs. 75th percentile).

---

## 3. Preparing Your Data

### What you need

causalGBKMR expects three inputs:

| Input | Type | Dimensions | Description |
|-------|------|-----------|-------------|
| **Y** | numeric vector | length $n$ | Outcome (continuous or binary) |
| **Z** | numeric matrix | $n \times (p \times T)$ | Exposures: $p$ mixture components at $T$ time points |
| **X** | numeric matrix | $n \times (L \times T + B)$ | $L$ time-dependent covariates at $T$ time points, plus $B$ baseline covariates |

### How to organize your columns

**Exposure matrix Z** --- group by time point, then by component:

```
[Mix1_t0, Mix2_t0, ..., Mixp_t0, Mix1_t1, ..., Mixp_t(T-1)]
```

**Covariate matrix X** --- time-dependent covariates first, then baseline:

```
[TDcov1_t1, TDcov2_t1, ..., TDcovL_t1, ..., TDcovL_t(T-1), Baseline1, ..., BaselineB]
```

> **Important:** Exposures (Z) include baseline $t=0$ through $t=T-1$.
> Time-dependent covariates (X) start at $t=1$ --- they are measured *after*
> baseline exposure. Baseline covariates (e.g., sex, age) go at the end of X.

### Data preparation tips

Before running causalGBKMR, we recommend the following checks:

```{r data-tips, eval=FALSE}
# 1. Check for missing data -- g-BKMR can handle some NAs but complete
#    cases are preferred
cat("Missing in Y:", sum(is.na(Y)), "\n")
cat("Missing in Z:", sum(is.na(Z)), "\n")
cat("Missing in X:", sum(is.na(X)), "\n")

# 2. Check exposure distributions -- environmental data is often right-skewed.
#    The package log-transforms by default (log_transform_mixtures = TRUE).
#    If your data is already log-transformed, set log_transform_mixtures = FALSE.
apply(Z, 2, summary)

# 3. Check for zero or negative values in exposures -- log(0) is undefined.
#    The package handles this by adding a small constant, but it's better to
#    address it beforehand (e.g., replace with LOD/sqrt(2)).
cat("Zeros in Z:", sum(Z == 0, na.rm = TRUE), "\n")
cat("Negatives in Z:", sum(Z < 0, na.rm = TRUE), "\n")

# 4. Check correlations between exposures -- high correlation is expected
#    (and handled by BKMR), but extreme collinearity (r > 0.99) may cause
#    numerical issues.
cor_mat <- cor(Z, use = "complete.obs")
cat("Max pairwise correlation:", max(abs(cor_mat[upper.tri(cor_mat)])), "\n")

# 5. Check sample size relative to model complexity.
#    Rule of thumb: you need at least 10 observations per predictor in the
#    largest BKMR model (the outcome model, which has p*T + L*(T-1) Z columns).
n_predictors_outcome <- ncol(Z) + ncol(X) - n_baseline
cat("Outcome model has", n_predictors_outcome, "predictors\n")
cat("Suggested minimum n:", 10 * n_predictors_outcome, "\n")
```

### Converting from common data formats

If your data is in long format (one row per person-visit), you need to reshape
to wide format first:

```{r long-to-wide, eval=FALSE}
library(tidyr)

# Example: long-format data with columns: id, visit, arsenic, lead, bmi, sex, outcome
wide_data <- long_data %>%
  pivot_wider(
    id_cols = c(id, sex, outcome),
    names_from = visit,
    values_from = c(arsenic, lead, bmi),
    names_sep = "_t"
  )

# Then extract matrices
Z <- as.matrix(wide_data[, c("arsenic_t0", "lead_t0", "arsenic_t1", "lead_t1")])
X <- cbind(
  as.matrix(wide_data[, c("bmi_t1")]),   # TD covariates (start at t1)
  wide_data$sex                            # baseline
)
Y <- wide_data$outcome
```

---

## 4. Scenario 1: Basic Analysis (Standard BKMR)

This is the standard use case for moderate sample sizes ($n < 2000$).

### Step 1: Generate example data

```{r scenario1-data}
set.seed(123)
n <- 200
T_points <- 3   # 3 time points (t=0, t=1, t=2)
p <- 2           # 2 exposures (e.g., arsenic, lead)
Ldim <- 1        # 1 time-dependent confounder (e.g., waist circumference)
n_baseline <- 1  # 1 baseline covariate (e.g., sex)

# Outcome
Y <- rnorm(n, mean = 50, sd = 10)

# Exposure matrix: 2 metals x 3 time points = 6 columns
# Using lognormal to mimic environmental concentrations
Z <- matrix(rlnorm(n * p * T_points, meanlog = 1, sdlog = 0.5),
            nrow = n, ncol = p * T_points)

# Covariate matrix: 1 TD covariate x 3 time points + 1 baseline = 4 columns
waist_base <- rnorm(n, 85, 12)
X <- cbind(
  waist_base + rnorm(n, 1, 2),    # waist at t=1
  waist_base + rnorm(n, 2, 2),    # waist at t=2
  waist_base + rnorm(n, 3, 2),    # waist at t=3
  rbinom(n, 1, 0.5)               # sex (baseline)
)

cat("Y:", length(Y), "observations\n")
cat("Z:", nrow(Z), "x", ncol(Z), "(", p, "exposures x", T_points, "time points)\n")
cat("X:", nrow(X), "x", ncol(X), "(", Ldim, "TD cov x", T_points, "times +",
    n_baseline, "baseline)\n")
```

### Step 2: Prepare data with `prepare_gbkmr_data()`

This function converts your matrices into the internal data frame format. It
handles variable naming, optional log-transformation, and input validation.

```{r scenario1-prepare}
prepared_data <- prepare_gbkmr_data(
  Y = Y,
  Z = Z,
  X = X,
  time_points = T_points,
  mixture_components = p,
  td_covariates = Ldim,
  baseline_covariates = n_baseline,
  td_covariate_names = "waist",        # give meaningful names to TD covariates
  log_transform_mixtures = TRUE         # log-transform (recommended for raw concentrations)
)

# Inspect the result
head(prepared_data[, 1:8])
```

The output data frame has columns in this order: baseline covariates, baseline
TD covariates, exposures at each time point, TD covariates at each time point,
outcome, and subject id.

### Step 3: Verify variable detection

Before running the analysis, check that the package correctly detected your
variable structure:

```{r scenario1-detect}
detection <- detect_variable_patterns(prepared_data, T = T_points)
print(detection)
```

This should show:

- `p` = number of exposures per time point
- `Ldim` = number of time-dependent covariates per time point
- The base names of your TD covariates

If the detection is wrong, check your column naming or provide explicit
parameters to `run_gbkmr_panel()` instead of using `gbkmr_run()`.

### Step 4: Run g-BKMR analysis

**For real analyses**, use production-quality settings:

```{r scenario1-run-production, eval=FALSE}
results <- gbkmr_run(
  data = prepared_data,
  outcome = "Y",
  outcome_type = "continuous",
  time_points = T_points,
  iter = 15000,                        # MCMC iterations (at least 10000)
  n = min(500, nrow(prepared_data)),   # subsample for BKMR fitting
  K = 1000,                            # Monte Carlo samples for g-computation
  n_knots = 50,                        # kernel approximation knots
  verbose = TRUE
)
# engine defaults to "auto": uses standard BKMR for n <= 2000,
# fast BKMR for n > 2000 (if fbkmr is installed).
# You can override with engine = "bkmr" or engine = "fastbkmr".

print(results)
summary(results)
```

For this vignette, we use small parameters so it finishes in seconds:

```{r scenario1-run-tiny}
results <- gbkmr_run(
  data = prepared_data,
  outcome = "Y",
  outcome_type = "continuous",
  time_points = T_points,
  iter = 500,       # tiny for demo only!
  n = 50,
  K = 10,
  n_knots = 15,
  verbose = TRUE
)
```

### Step 5: Interpret results

The `results` object is a `gbkmr_results` list with several components.
The simplest way to view the headline finding is to print it:

```{r scenario1-print}
print(results)
```

For a more detailed summary including settings:

```{r scenario1-summary}
summary(results)
```

**How to read the output:**

- **ATE (Average Treatment Effect):** the causal contrast between the high and
  low exposure scenarios. Here, `r round(results$causal_effect$estimate, 3)`
  means that shifting all mixture components from their 25th to 75th percentile
  is associated with a causal change of approximately this magnitude in the
  outcome.
- **95% Credible Interval:** the Bayesian analogue of a confidence interval.
  If this interval excludes zero (same sign for lower and upper bounds), we
  conclude the exposure contrast has a non-null causal effect at the 5% level.
- **Counterfactual means:** `E[Y^a]` and `E[Y^a*]` are the expected outcomes
  under each exposure scenario, averaged over the distribution of confounders.
  The ATE is their difference.

You can also access individual components directly:

```{r scenario1-components}
# Causal effect estimate and CI
results$causal_effect

# Counterfactual outcome means
results$counterfactual_means

# Posterior inclusion probabilities (variable importance from outcome model)
# These are beta coefficients from the linear part; for PIPs on mixture
# components, see results$raw_results$fit_y
round(results$variable_importance, 4)
```

**Interpreting significance:** In this tiny-sample demo, the credible interval
likely includes zero (we're using only 50 subjects and 500 MCMC iterations).
In a real analysis with production settings, a CI that excludes zero would
constitute evidence of a causal effect of the mixture on the outcome.

### Step 6: Diagnostic plots

Always check convergence before trusting results.

```{r scenario1-diagnostics, fig.width=7, fig.height=4}
raw <- results$raw_results

# Posterior distribution of ATE
diff_posterior <- raw$Yastar - raw$Ya
hist(diff_posterior, breaks = 30, col = "steelblue", border = "white",
     main = "Posterior Distribution of ATE",
     xlab = "ATE (high - low exposure)")
abline(v = results$causal_effect$estimate, col = "red", lwd = 2)
abline(v = 0, col = "gray40", lty = 2, lwd = 2)
legend("topright", c("Estimate", "Zero"), col = c("red", "gray40"),
       lty = c(1, 2), lwd = 2, bty = "n")
```

```{r scenario1-trace, fig.width=7, fig.height=3.5}
# Trace plots for convergence
par(mfrow = c(1, 2))
plot(raw$Ya, type = "l", col = "steelblue",
     main = "E[Y | low exposure]", ylab = "Y(a)", xlab = "MCMC iteration")
plot(raw$Yastar, type = "l", col = "tomato",
     main = "E[Y | high exposure]", ylab = "Y(a*)", xlab = "MCMC iteration")
par(mfrow = c(1, 1))
```

**What to look for:** Trace plots should look like "hairy caterpillars" ---
stationary with no trends. If you see drift or slow mixing, increase `iter`.

---

## 5. Scenario 2: Custom Intervention Levels

By default, g-BKMR compares all exposures at the 25th percentile (low) vs.
75th percentile (high). You can customize this in two ways.

### Option A: Change the quantile probabilities

```{r scenario2-quantiles, eval=FALSE}
# Compare 10th vs 90th percentile (wider contrast)
results_wide <- gbkmr_run(
  data = prepared_data,
  outcome = "Y",
  outcome_type = "continuous",
  time_points = T_points,
  a_probs = c(0.10, 0.90),      # 10th vs 90th percentile
  iter = 15000, n = 500, K = 1000
)
```

### Option B: Specify exact exposure values

Use this when you want to compare specific exposure scenarios (e.g., a
regulatory standard vs. observed levels). You need `run_gbkmr_panel()` for this:

```{r scenario2-custom, eval=FALSE}
# Build named vectors matching exposure column names
exposure_names <- paste0("logM", rep(1:2, 3), "_", rep(0:2, each = 2))

# Low scenario: all exposures at log(1) (i.e., concentration = 1)
a_low <- setNames(rep(0, length(exposure_names)), exposure_names)

# High scenario: all exposures at log(5) (i.e., concentration ~ 5)
a_high <- setNames(rep(log(5), length(exposure_names)), exposure_names)

results_custom <- run_gbkmr_panel(
  sim_popn = prepared_data,
  T = T_points,
  p = 2,
  mediator_basenames = "waist",
  common_covariates = c("sex", "waist_0"),
  a_vals = a_low,
  astar_vals = a_high,
  iter = 15000, n = 500, K = 1000, n_knots = 50
)

cat("ATE:", results_custom$diff_gBKMR, "\n")
```

---

## 6. Scenario 3: Large-Sample Analysis with fastBKMR

When sample size is large ($n > 2000$), standard BKMR becomes computationally
expensive because the Gaussian process kernel scales as $O(n^3)$. **fastBKMR**
(via the `fbkmr` package) addresses this by splitting data into subsets, fitting
BKMR on each in parallel, and combining results.

### When to use fastBKMR

The package selects the engine **automatically** (`engine = "auto"`, the default):

| Sample size | Auto behavior |
|-------------|--------------|
| $n \le 2000$ | Standard BKMR with knot approximation |
| $n > 2000$ (fbkmr installed) | fastBKMR, with `n_subset = max(5, floor(n/1000))` |
| $n > 2000$ (fbkmr not installed) | Standard BKMR + message suggesting fbkmr |

**Why 2000?** The fastBKMR authors recommend ~1000 observations per subset
(Sonabend et al., 2024). With at least 2 subsets needed, the minimum useful
sample size is 2000. Below that, standard BKMR with knot approximation
remains tractable (~80 min/model at n=2000 in our benchmarks).

You can always override the auto-selection:

### Install fastBKMR

```{r scenario3-install, eval=FALSE}
# fastBKMR is not on CRAN -- install from GitHub
remotes::install_github("junwei-lu/fbkmr")

# Verify
library(fbkmr)
?skmbayes
```

### Running with fastBKMR

The interface is **identical** to standard BKMR --- just change `engine`:

```{r scenario3-run, eval=FALSE}
# Suppose you have a large prepared dataset (n = 6000)
results_fast <- gbkmr_run(
  data = large_prepared_data,
  outcome = "Y",
  outcome_type = "continuous",
  time_points = 3,
  engine = "fastbkmr",           # or omit -- auto-selected when n > 2000
  # n_subset and n_cores are auto-calculated when NULL:
  # n_subset = max(5, floor(n/1000)), n_cores = min(n_subset, cores, 10)
  iter = 15000,
  n = nrow(large_prepared_data), # use all data (fastBKMR handles the scaling)
  K = 1000,
  verbose = TRUE
)

print(results_fast)
```

### Key parameters for fastBKMR

| Parameter | Default | Description |
|-----------|---------|-------------|
| `engine` | `"auto"` | `"auto"` selects based on n; or force `"bkmr"` / `"fastbkmr"` |
| `n_subset` | `10` | Number of data subsets. More subsets = faster per-subset fitting, but slightly more variance |
| `n_cores` | `10` | Parallel cores. Set to `n_subset` for maximum parallelism |

**Tip on choosing `n_subset`:** Each subset should have at least ~200
observations. So for $n = 6000$, `n_subset = 10` gives ~600 per subset
(good). For $n = 2000$, try `n_subset = 5` to keep ~400 per subset.

### How it works under the hood

1. `fbkmr::skmbayes()` splits the data into `n_subset` random subsets
2. Standard BKMR (`bkmr::kmbayes()`) is fit independently on each subset, in
   parallel across `n_cores` cores
3. During g-computation, the package calls `bkmr::SamplePred()` on a
   **randomly selected subset** at each MCMC iteration
4. Results (betas, predictions) are aggregated across subsets

> **Technical note:** `bkmr::SamplePred()` works on individual subset fits
> (class `bkmrfit`) but fails on the full `skmbayes` list object. causalGBKMR
> handles this automatically --- no user action needed.

### Advanced: using `run_gbkmr_panel()` directly

For full control over model fitting, you can bypass `gbkmr_run()` and call the
core function:

```{r scenario3-panel, eval=FALSE}
results_fast <- run_gbkmr_panel(
  sim_popn = large_prepared_data,
  T = 3,
  p = 4,
  mediator_basenames = c("waist", "bmi"),
  common_covariates = c("sex", "waist_0", "bmi_0"),
  engine = "fastbkmr",
  n_subset = 10,
  n_cores = 10,
  n = 6000,
  iter = 15000,
  sel = seq(12000, 15000, by = 25),   # post-burn-in samples
  K = 1000,
  verbose_every = 20
)
```

---

## 7. Scenario 4: Working with Mixed Variable Types

Real studies often mix continuous and binary variables. The package
auto-detects variable types from your data and prints what it understood
in the **Input Audit** block at the top of `gbkmr_run()` output.

### Example: binary outcome, binary TD confounder

```{r scenario4-setup, eval=FALSE}
set.seed(7)
n <- 300
T_points <- 3

# Binary outcome (e.g., incident diabetes by visit 3)
Y <- rbinom(n, 1, 0.3)

# Mixture: 2 continuous exposures at 3 time points
Z <- matrix(rlnorm(n * 6, 1, 0.5), n, 6)

# TD confounder: smoking status (binary), measured at t=1 and t=2,
# plus baseline smoking, plus sex (baseline)
X <- cbind(
  rbinom(n, 1, 0.25),  # smoking_1 (binary TD)
  rbinom(n, 1, 0.27),  # smoking_2 (binary TD)
  rbinom(n, 1, 0.30),  # smoking_3 (binary TD)
  rbinom(n, 1, 0.5)    # sex (baseline)
)

prepared <- prepare_gbkmr_data(
  Y = Y, Z = Z, X = X,
  time_points = T_points,
  mixture_components = 2,
  td_covariates = 1,
  baseline_covariates = 1,
  td_covariate_names = "smoking"
)

results_mixed <- gbkmr_run(
  data = prepared,
  outcome_type = "binary",      # <-- tell the package it's binary
  time_points = T_points,
  iter = 15000
)
```

**What to expect in the Input Audit output:**

```
[Outcome]
  Variable:        Y
  Detected type:   binary
  User specified:  binary
  Link function:   probit (via bkmr::kmbayes family='binomial')
  NOTE: binary outcome support is experimental in this version.

[Time-varying confounders]
  Number per time (Ldim):  1
  Names:                   smoking
    - smoking             type: binary
      NOTE: binary TD confounder -- BKMR mediator model
            will treat it as continuous in current version.
```

> **Current limitation:** Binary TD confounders are fit with Gaussian BKMR
> in this version (not probit). The mediator sampling step still produces
> valid g-computation, but posterior coverage may be optimistic. Full
> binary-TD support is planned for v2.

### What the package auto-detects vs.\ what you must specify

| Variable role | Auto-detected? | Where to specify if needed |
|---------------|---------------|---------------------------|
| Outcome type (continuous/binary) | Yes (warns if mismatch) | `outcome_type` argument |
| TD confounder type | Yes (displayed in audit) | *(not yet configurable)* |
| Mixture components | Yes (from column naming) | `mixture_components` in prep |
| Number of time points | No --- user input | `time_points` argument |
| Baseline covariates | Yes (from column position) | `baseline_covariates` in prep |

---

## 8. Scenario 5: Setting Custom Contrasts Correctly

This is the most commonly misunderstood part of causal mixture analysis.
The causal contrast---comparing exposure scenario $\bar{a}$ versus
$\bar{a}^*$---**must only change the mixture components**, never the
confounders or baseline covariates.

### The default: all mixture components at 25th vs.\ 75th percentile

```{r scenario5-default, eval=FALSE}
# Same as omitting a_probs (default is c(0.25, 0.75))
results <- gbkmr_run(
  data = prepared_data, time_points = 3,
  a_probs = c(0.25, 0.75)
)
```

The Input Audit will print:

```
[Causal contrast]
  Low exposure (a):   all mixture components at 25th percentile
  High exposure (a*): all mixture components at 75th percentile
  IMPORTANT: Only mixture columns vary between a and a*.
             Time-varying confounders are sampled via g-computation.
             Baseline covariates are held at their mean.
```

### Wider or narrower contrasts

```{r scenario5-widecontrast, eval=FALSE}
# 10th vs 90th percentile -- stronger contrast, wider CI
results_wide <- gbkmr_run(
  data = prepared_data, time_points = 3,
  a_probs = c(0.10, 0.90)
)
```

### Specific exposure values (e.g., regulatory thresholds)

Suppose you want to compare "all metals at background levels (1 ug/L)"
versus "all metals at the EPA action level (5 ug/L)". Because the
package log-transforms mixtures by default, you need to pass `log(1) = 0`
and `log(5) = 1.609` in the *transformed* scale:

```{r scenario5-custom, eval=FALSE}
# Get the exposure column names first
exposure_cols <- paste0("logM", rep(1:2, 3), "_", rep(0:2, each = 2))
# exposure_cols = c("logM1_0", "logM2_0", "logM1_1", "logM2_1", "logM1_2", "logM2_2")

a_background <- setNames(rep(log(1), 6), exposure_cols)  # background = 1 ug/L
a_action     <- setNames(rep(log(5), 6), exposure_cols)  # action level = 5 ug/L

results_regulatory <- run_gbkmr_panel(
  sim_popn = prepared_data,
  T = 3, p = 2,
  mediator_basenames = "waist",
  common_covariates = c("sex", "waist_0"),
  a_vals = a_background,
  astar_vals = a_action,
  iter = 15000, n = 500, K = 1000, n_knots = 50
)
```

### What NOT to do: accidentally varying confounders in the contrast

A common mistake when using `run_gbkmr_panel()` directly is to put
non-mixture variables in `a_vals` / `astar_vals`:

```{r scenario5-wrong, eval=FALSE}
# WRONG -- don't do this
a_wrong <- c(logM1_0 = -1, logM2_0 = -1, waist_0 = 80)  # <-- waist shouldn't be here!
```

**Why it's wrong:** `waist_0` is a baseline confounder, not an exposure.
Changing it between scenarios would mean you're estimating the effect of
"different confounder values," which is not a causal mixture effect.

The Input Audit always prints which columns will vary (mixture only),
so double-check that block before launching a long MCMC run.

---

## 9. Scenario 6: Convergence Diagnostics Workflow

MCMC convergence is **not guaranteed** just because the function
completes. Always check diagnostics before trusting results.

### Quick visual check: posterior stability

```{r scenario6-diag1, eval=FALSE}
raw <- results$raw_results

# Posterior draws of E[Y|a] and E[Y|a*] should be "hairy caterpillars"
# -- stationary with no drift
par(mfrow = c(1, 2))
plot(raw$Ya, type = "l", col = "steelblue",
     main = "E[Y | low exposure]", ylab = "Y(a)", xlab = "MCMC iteration")
plot(raw$Yastar, type = "l", col = "tomato",
     main = "E[Y | high exposure]", ylab = "Y(a*)", xlab = "MCMC iteration")
par(mfrow = c(1, 1))
```

**What to look for:**

- Stable horizontal band → converged
- Obvious drift or slow trend → increase `iter` and/or `sel` burn-in
- Sudden jumps → examine the DGP; may indicate multimodal posterior

### Diagnosing individual BKMR models

Each fitted mediator and outcome model is stored in `results$raw_results`:

```{r scenario6-diag2, eval=FALSE}
# Mediator models (one per time-varying confounder per visit)
fit_mediators <- results$raw_results$fit_mediators
# fit_mediators[[t]][[l]] is the BKMR fit for TD confounder l at time t

# Outcome model
fit_y <- results$raw_results$fit_y

# For standard BKMR, use the bkmr diagnostic functions directly:
if (inherits(fit_y, "bkmrfit")) {
  bkmr::TracePlot(fit = fit_y, par = "beta")
  bkmr::TracePlot(fit = fit_y, par = "sigsq.eps")
}

# For fastBKMR, diagnostics are per-subset:
if (is.list(fit_y) && !inherits(fit_y, "bkmrfit")) {
  # Each fit_y[[i]] is a bkmrfit from subset i
  bkmr::TracePlot(fit = fit_y[[1]], par = "beta")
}
```

### If convergence looks poor

Try these, in order:

1. **Increase `iter`** from 15,000 to 30,000 or 50,000
2. **Increase burn-in**: `sel = seq(floor(iter * 0.8), iter, by = 25)`
3. **Check for problematic variables**: extreme outliers, near-constant
   columns, or highly collinear exposures can destabilize MCMC
4. **Run multiple chains** (currently requires manual loop over `currind`
   seeds; planned for v2)

> **Warning about packaged workflows:** when running `gbkmr_run()` with
> many time points and confounders, the package fits many BKMR models
> internally. A single non-converging model can bias the causal estimate.
> In the current version, we recommend checking each mediator model
> individually for studies with $T > 3$ or complex exposure structures.

---

## 10. Tips and Best Practices

### Parameter tuning guide

| Goal | Adjustment |
|------|-----------|
| **Faster prototyping** | Reduce `iter` (5000), `K` (200), `n` (200) |
| **More stable estimates** | Increase `K` (2000) and `length(sel)` |
| **Memory issues** | Reduce `n_knots` (30) or `n` |
| **Large dataset** | Automatic when n > 2000, or `engine = "fastbkmr"` |
| **Binary outcome** | Set `outcome_type = "binary"` |

### Recommended production settings

```{r production, eval=FALSE}
# For any sample size -- engine is auto-selected:
results <- gbkmr_run(
  data = prepared_data,
  time_points = T_points,
  iter = 20000,
  n = min(800, nrow(prepared_data)),
  K = 1000,
  n_knots = 50
)
# If n <= 2000: uses standard BKMR
# If n >  2000: uses fastBKMR (auto n_subset, n_cores)

# Or force a specific engine:
results <- gbkmr_run(
  data = prepared_data,
  time_points = T_points,
  iter = 15000,
  n = nrow(prepared_data),
  K = 1000,
  engine = "fastbkmr"     # override auto-selection
)
```

### Common pitfalls

1. **Forgetting to log-transform exposures.** Environmental concentrations are
   typically right-skewed. `prepare_gbkmr_data()` sets
   `log_transform_mixtures = TRUE` by default. If your data is *already*
   log-transformed, set it to `FALSE` to avoid double-transforming.

2. **Too few MCMC iterations.** With `iter = 5000`, BKMR may not converge.
   Use at least `iter = 15000` for real analyses. Always check trace plots
   (see Scenario 1, Step 6).

3. **`sel` exceeds `iter`.** If you set `sel = seq(22000, 24000, ...)` but
   `iter = 15000`, you will get an error. The package auto-calculates `sel`
   when using `gbkmr_run()`, but with `run_gbkmr_panel()` you must ensure
   `max(sel) <= iter`.

4. **Knot warnings.** Messages like "Number of nearest neighbors reduced" from
   `fields::cover.design()` are harmless when $n$ is small relative to
   `n_knots`. Reduce `n_knots` if they bother you.

5. **Columns below the limit of detection (LOD).** For values below LOD,
   replace with `LOD / sqrt(2)` before passing to causalGBKMR. Do not leave
   as zero (log(0) is undefined).

---

## 11. Function Reference

| Function | Purpose | When to use |
|----------|---------|-------------|
| `prepare_gbkmr_data()` | Convert matrices (Y, Z, X) to g-BKMR format | Always --- first step |
| `validate_user_matrices()` | Check input dimensions and types | Debugging data issues |
| `detect_variable_patterns()` | Auto-detect variable structure | Before running, to verify setup |
| `gbkmr_run()` | High-level analysis wrapper | Recommended for most users |
| `run_gbkmr_panel()` | Core g-BKMR with full parameter control | Advanced: custom interventions, fine-tuning |

---

## Session Info

```{r session-info}
sessionInfo()
```
