---
title: "Diagnostics & Troubleshooting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gbkmr_diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

g-BKMR fits many BKMR models under the hood (one per time-varying
confounder per visit, plus the outcome model). Any one of them can fail
to converge, which biases the causal estimate. This article shows how
to detect and fix problems.

## 1. Input Audit (before MCMC)

Every call to `gbkmr_run()` prints an **Input Audit** block before MCMC
launches. Verify each section before letting a long run proceed.

```
==============================================================
  causalBKMR: Input Audit
==============================================================

[Outcome]
  Variable:        Y
  Detected type:   continuous
  User specified:  continuous
  Link function:   identity (Gaussian BKMR)

[Mixture exposures]
  Components per time (p): 2
  Time points (T):         3
  Total exposure columns:  6

[Time-varying confounders]
  Number per time (Ldim):  1
  Names:                   waist
    - waist                type: continuous

[Baseline covariates]
  Variables: sex, waist_0
  Role:      linear term X*beta (NOT in kernel h())

[Model configuration]
  Engine:          bkmr
  Knots:           50
  Sample size:     500
  MCMC iterations: 15000
  MC samples (K):  1000

[Causal contrast]
  Low exposure (a):   all mixture components at 25th percentile
  High exposure (a*): all mixture components at 75th percentile
```

If any of these is unexpected — stop. Common causes:

- Outcome variable mismatch (e.g., `Y` is binary 0/1 but you specified
  `outcome_type = "continuous"`).
- Wrong column count (e.g., you forgot a baseline covariate).
- TD confounder type unexpected.

## 2. Automatic convergence check (after MCMC)

When `verbose = TRUE` (default), the package prints a `[Convergence
check]` block after MCMC finishes:

```
[Convergence check]
  [OK]      mediator_t1_1        ESS=423, Geweke |z|=0.85
  [OK]      mediator_t2_1        ESS=389, Geweke |z|=1.20
  [WARNING] outcome_Y            ESS=67, Geweke |z|=3.20  [low_ess, geweke_nonstationary]

  Some models show signs of poor convergence. Consider:
    - Increasing iter (try 30000 or 50000)
    - Increasing burn-in: sel = seq(floor(iter*0.8), iter, by=25)
    - Inspecting trace plots: bkmr::TracePlot(results$raw_results$fit_y)
```

What the checks mean:

- **ESS (Effective Sample Size).** Lower bound on the equivalent number
  of independent posterior draws. We flag `low_ess` when ESS < 100 for
  any parameter.
- **Geweke z-score.** Compares chain means in the first 10% to the last
  50% of post-burn-in draws. We flag `geweke_nonstationary` when
  $|z| > 2.5$.

The diagnostics are also stored on the results object:

```{r diag-obj}
results$diagnostics
# $mediator_t1_1
# $ess = 423
# $geweke_max_abs = 0.85
# $warning_flags = character(0)
# ...
```

## 3. Visual diagnostics

For deeper inspection, use the underlying `bkmr` diagnostic functions
directly on the stored fits:

### Standard BKMR engine

```{r vis-bkmr}
library(bkmr)
fit_y <- results$raw_results$fit_y

# Trace plot of beta (linear covariate coefficient)
TracePlot(fit = fit_y, par = "beta")

# Trace plot of residual variance
TracePlot(fit = fit_y, par = "sigsq.eps")

# Trace plot of kernel parameter
TracePlot(fit = fit_y, par = "r")
```

What to look for:

- **Stable horizontal band** ("hairy caterpillar"). Good.
- **Drift or trend** over iterations. Increase `iter`.
- **Stuck plateau** for many iterations. Multi-modal posterior;
  consider reparameterisation or different priors.

### fastBKMR engine

For fastBKMR fits, each subset has its own trace:

```{r vis-fast}
# results$raw_results$fit_y is a list of bkmrfit objects
fit_list <- results$raw_results$fit_y

# Plot subset 1
bkmr::TracePlot(fit = fit_list[[1]], par = "beta")

# Or loop over all subsets
par(mfrow = c(2, 5))   # 10 subsets
for (i in seq_along(fit_list)) {
  bkmr::TracePlot(fit = fit_list[[i]], par = "beta")
}
```

## 4. Sampling diagnostics for the causal effect

The g-computation draws produce posterior samples of the **causal
effect** itself. Plot the trace to check posterior stability:

```{r vis-ace}
raw <- results$raw_results

par(mfrow = c(1, 2))
plot(raw$Ya, type = "l", col = "steelblue",
     main = "E[Y | low exposure]",
     ylab = expression(Y(bar(a))), xlab = "MCMC iteration")
plot(raw$Yastar, type = "l", col = "tomato",
     main = "E[Y | high exposure]",
     ylab = expression(Y(bar(a)^"*")), xlab = "MCMC iteration")

# Posterior distribution of the ACE itself
diff_post <- raw$Yastar - raw$Ya
hist(diff_post, breaks = 30, col = "steelblue", border = "white",
     main = "Posterior of ACE", xlab = "ACE")
abline(v = 0, col = "gray40", lty = 2, lwd = 2)
abline(v = results$causal_effect$estimate, col = "red", lwd = 2)
```

## 5. Common problems and fixes

| Symptom | Likely cause | Fix |
|---------|-------------|-----|
| `low_ess` on `outcome_Y` | Too few MCMC iterations | Increase `iter` to 30,000+ |
| `geweke_nonstationary` | Chain hasn't reached stationarity | Increase burn-in: `sel = seq(floor(iter*0.8), iter, by=25)` |
| Cf means outside reasonable range | Outcome on wrong scale / unintended transformation | Check `log_transform_mixtures` in `prepare_gbkmr_data()` |
| Cf means not in [0,1] for binary | `outcome_type` mis-specified | Set `outcome_type = "binary"` |
| `Number of nearest neighbors reduced` warning | `n_knots` > unique rows | Reduce `n_knots`; harmless otherwise |
| `Parallel execution failed` warning | login node restricts parallelism | Package auto-falls back to sequential; submit job to a compute node for speed |
| Wildly different PIPs across runs | Pre-burn-in samples included | Verify `max(sel) <= iter` and `sel` skips burn-in |

## 6. Best-practice settings

For a real analysis (not a smoke test):

```{r prod}
results <- gbkmr_run(
  data = prepared_data,
  time_points = 3,
  iter = 30000,         # at least 15K, 30K+ for safety
  K = 1000,             # Monte Carlo samples for g-computation
  n_knots = 50          # standard BKMR only
  # engine defaults to "auto", n_subset/n_cores auto-set for fastBKMR
)

# If any [WARNING] in convergence check, rerun with:
results <- gbkmr_run(
  data = prepared_data,
  time_points = 3,
  iter = 60000,
  sel = seq(floor(60000 * 0.8), 60000, by = 25)
)
```

## See also

- [Methodology overview](gbkmr_method_overview.html)
- [Quick Start](gbkmr_quickstart.html)
- `?bkmr::TracePlot` — the underlying trace-plot function
