---
title: "Example 2: Binary outcome (probit BKMR)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gbkmr_example_binary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This example shows how to run g-BKMR with a **binary outcome** (e.g.,
incident diabetes, hospitalisation). Under the hood the package fits
a **probit BKMR** (`family = "binomial"` in `bkmr::kmbayes`) for the
outcome model and converts predictions back to the probability scale.
The ACE then represents a **causal risk difference**.

## Setup

```{r setup-data}
library(causalBKMR)

set.seed(7)
n <- 300
T_points <- 3

# Binary outcome (e.g., incident diabetes by visit 3)
Y <- rbinom(n, 1, 0.3)

# Mixture: 2 continuous exposures at 3 visits
Z <- matrix(rlnorm(n * 6, 1, 0.5), n, 6)

# 1 TD confounder (waist circumference) measured at 3 visits + 1 baseline (sex)
waist_base <- rnorm(n, 85, 12)
X <- cbind(
  waist_base + rnorm(n, 1, 2),    # waist at t=1
  waist_base + rnorm(n, 2, 2),    # waist at t=2
  waist_base + rnorm(n, 3, 2),    # waist at t=3
  rbinom(n, 1, 0.5)               # sex (baseline)
)

prepared <- prepare_gbkmr_data(
  Y = Y, Z = Z, X = X,
  time_points = T_points,
  mixture_components = 2,
  td_covariates = 1,
  baseline_covariates = 1,
  td_covariate_names = "waist"
)
```

## Run with `outcome_type = "binary"`

```{r run-binary}
results_binary <- gbkmr_run(
  data = prepared,
  outcome_type = "binary",      # <- triggers probit BKMR
  time_points = T_points,
  iter = 15000
)
```

## What changes for binary outcomes

The **Input Audit** flags binary status:

```
[Outcome]
  Variable:        Y
  Detected type:   binary
  User specified:  binary
  Link function:   probit (via bkmr::kmbayes family='binomial')
  NOTE: Only engine='bkmr' supports binary outcomes.
        Output is returned on the probability scale.
```

Under the hood:

1. **Outcome model** uses `bkmr::kmbayes(..., family = "binomial")` ->
   fits probit BKMR.
2. After MCMC, `bkmr::SamplePred()` returns the linear predictor
   $h(\mathbf{Z}) + \mathbf{X}^\top\boldsymbol{\beta}$. The package
   transforms via $\Phi(\cdot)$ (standard normal CDF) to get
   probabilities.
3. ACE is therefore a **risk difference**:
   $\text{ACE} = P(Y=1 \mid \bar{\mathbf{a}}^*) - P(Y=1 \mid \bar{\mathbf{a}})$.

## Read the output

```{r read-binary}
print(results_binary)

# ACE is a risk difference; can be positive or negative
results_binary$causal_effect$estimate   # e.g., 0.04 = a 4 percentage-point increase

# Counterfactual probabilities (both should be in [0, 1])
results_binary$counterfactual_means$low   # P(Y=1 | a)
results_binary$counterfactual_means$high  # P(Y=1 | a*)
```

## Limitations

### Sample size

In v1, binary-outcome g-BKMR only uses the **standard BKMR engine**.
We recommend $n \le 1{,}500$ for binary outcomes; larger samples may
need scaling tricks. fastBKMR with binary outcome is on the v2 roadmap
(waiting on a polygamma sampler in upstream `fbkmr`).

```{r fastbkmr-binary, eval = FALSE}
# This will ERROR with a clear message:
gbkmr_run(data = prepared, outcome_type = "binary",
          engine = "fastbkmr")
#> Error: Binary outcome is not supported with engine='fastbkmr'.
#>   fbkmr::skmbayes() does not expose the family argument.
#>   Use engine='bkmr' for binary outcomes, or subsample your data.
```

### Binary time-varying confounders

Currently, if you have a **binary** time-varying confounder (e.g.,
smoking status that varies over visits), the package fits the mediator
model with Gaussian BKMR and the Input Audit warns:

```
[Time-varying confounders]
    - smoking             type: binary
      NOTE: binary TD confounder -- BKMR mediator model
            will treat it as continuous in current version.
```

The mediator sampling step uses normal noise rather than Bernoulli. The
g-computation predictions are still defined but posterior coverage may
be slightly optimistic for the binary confounder pathway. Full probit
support for binary $L_t$ is planned for v2.

## When to use binary outcomes

- Outcome is rare disease incidence (e.g., new diabetes case)
- Outcome is yes/no event during follow-up
- Outcome is a thresholded clinical indicator

For **continuous** outcomes (log-glucose, BMI, blood pressure), use the
default `outcome_type = "continuous"` (Gaussian BKMR).

## See also

- [Methodology overview](gbkmr_method_overview.html)
- [Quick Start](gbkmr_quickstart.html)
- [Diagnostics & troubleshooting](gbkmr_diagnostics.html)
