---
title: "Quick Start: g-BKMR with a continuous outcome"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gbkmr_quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  eval = requireNamespace("causalBKMR", quietly = TRUE)
)
```

This vignette walks through a complete g-BKMR analysis on simulated
data: 2 mixture components measured at 3 time points, 1 time-varying
confounder, 1 baseline covariate, and a continuous outcome. For the
methodological background, see the
[Methodology Overview](gbkmr_method_overview.html).

## 1. Load the package

```{r load}
library(causalBKMR)
```

## 2. Simulate example data

```{r simdata}
set.seed(123)
n <- 200
T_points <- 3   # 3 visits: t = 0, 1, 2
p <- 2           # 2 exposures per visit
Ldim <- 1        # 1 time-varying confounder
n_baseline <- 1  # 1 baseline covariate (sex)

# Continuous outcome
Y <- rnorm(n, mean = 50, sd = 10)

# Exposure matrix: 2 metals x 3 visits = 6 columns
# Lognormal to mimic environmental concentrations
Z <- matrix(rlnorm(n * p * T_points, meanlog = 1, sdlog = 0.5),
            nrow = n, ncol = p * T_points)

# Covariate matrix: 1 TD covariate (waist) x 3 visits + 1 baseline (sex) = 4 cols
waist_base <- rnorm(n, 85, 12)
X <- cbind(
  waist_base + rnorm(n, 1, 2),    # waist at t=1
  waist_base + rnorm(n, 2, 2),    # waist at t=2
  waist_base + rnorm(n, 3, 2),    # waist at t=3
  rbinom(n, 1, 0.5)               # sex (baseline)
)

cat("Y:", length(Y), "obs\n")
cat("Z:", nrow(Z), "x", ncol(Z), "(", p, "exposures x", T_points, "visits)\n")
cat("X:", nrow(X), "x", ncol(X), "(", Ldim, "TD x", T_points, "+",
    n_baseline, "baseline)\n")
```

## 3. Prepare data

`prepare_gbkmr_data()` converts your three matrices into the internal
wide-format frame, naming columns as `logM<i>_<t>` for exposures and
`<name>_<t>` for time-varying confounders.

```{r prep}
prepared_data <- prepare_gbkmr_data(
  Y = Y, Z = Z, X = X,
  time_points = T_points,
  mixture_components = p,
  td_covariates = Ldim,
  baseline_covariates = n_baseline,
  td_covariate_names = "waist",
  log_transform_mixtures = TRUE   # recommended for raw concentrations
)

head(prepared_data[, 1:8])
```

## 4. Verify what the package detected

```{r detect}
detection <- detect_variable_patterns(prepared_data, T = T_points)
print(detection)
```

The detection report confirms:

- `p = 2` exposure components per visit
- `Ldim = 1` time-varying confounder named `waist`
- Naming pattern recognised as `known_with_underscore`

If anything here is unexpected, stop and check your matrix construction
before running MCMC.

## 5. Run the analysis

For production analyses, use `iter = 15000` or higher. For this vignette
we use tiny parameters so it finishes in seconds.

```{r run}
results <- gbkmr_run(
  data = prepared_data,
  outcome = "Y",
  outcome_type = "continuous",
  time_points = T_points,
  iter = 500,       # tiny demo; use 15000+ in practice
  n = 50,
  K = 10,
  n_knots = 15
  # engine = "auto"  # default: picks bkmr or fastbkmr based on n
)
```

Before MCMC starts, you will see an **Input Audit** block that shows
exactly what the package understood (outcome type, mixture dimensions,
TD confounder types, model configuration, intervention contrast). If
anything is wrong, abort and re-check.

## 6. Read the results

```{r print}
print(results)
```

`print()` shows the headline ACE and 95% credible interval, plus the
counterfactual means under low and high exposure.

```{r summary}
summary(results)
```

`summary()` adds the analysis settings.

### Interpreting the output

- **ACE.** The causal contrast on the **same scale as the outcome**.
  Here Y is continuous, so the ACE is the expected change in Y when
  shifting the mixture from the 25th to 75th percentile (under the
  observed confounder distribution).
- **95% credible interval.** Bayesian analogue of a CI. Excluding zero
  (lower and upper same sign) is evidence for a non-null effect.
- **Counterfactual means.** $E[Y^{a}]$ and $E[Y^{a^*}]$ averaged over
  confounders. Their difference equals the ACE.

```{r components}
results$causal_effect       # estimate, lower, upper
results$counterfactual_means
round(results$variable_importance, 4)  # linear-term beta from outcome model
```

## 7. Convergence diagnostics

Every call to `gbkmr_run()` runs Effective Sample Size (ESS) and Geweke
z-score checks on each fitted BKMR model and stores them in
`results$diagnostics`. In verbose mode the package also prints a
`[Convergence check]` block (look for `[OK]` or `[WARNING]` per model).

```{r diag}
results$diagnostics
```

If any model shows `low_ess` or `geweke_nonstationary`, increase `iter`,
increase burn-in (`sel`), and re-run.

For full diagnostic workflows including trace plots, see
[Diagnostics & Troubleshooting](gbkmr_diagnostics.html).

## What's next?

- [Example 1 — Custom intervention contrasts](gbkmr_example_contrasts.html)
- [Example 2 — Binary outcome](gbkmr_example_binary.html)
- [Example 3 — Large sample with fastBKMR](gbkmr_example_fastbkmr.html)
- [Diagnostics & troubleshooting](gbkmr_diagnostics.html)
