---
title: "Example 1: Custom intervention contrasts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gbkmr_example_contrasts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

The default g-BKMR contrast compares all mixture components at the
**25th vs. 75th percentile**. This is convenient but not always the
right scientific question. This article shows how to specify:

- Different quantile contrasts (e.g., 10th vs. 90th)
- Specific exposure values (e.g., regulatory thresholds)
- What you **should not** do

For all examples we assume `prepared_data` is the output of
`prepare_gbkmr_data()` with 2 mixture components and 3 visits.

## The default

```{r default}
library(causalBKMR)
results <- gbkmr_run(
  data = prepared_data,
  time_points = 3,
  a_probs = c(0.25, 0.75)   # implicit default
)
```

The Input Audit will print:

```
[Causal contrast]
  Low exposure (a):   all mixture components at 25th percentile
  High exposure (a*): all mixture components at 75th percentile
  Target estimand:    ACE = E[Y(a*)] - E[Y(a)]
  IMPORTANT: Only mixture columns vary between a and a*.
             Time-varying confounders are sampled via g-computation.
             Baseline covariates are held at their mean.
```

## Option A — different quantile pair

```{r wide}
# 10th vs 90th percentile -- stronger contrast, wider CI
results_wide <- gbkmr_run(
  data = prepared_data,
  time_points = 3,
  a_probs = c(0.10, 0.90),
  iter = 15000
)
```

## Option B — specific exposure values

Suppose you want to compare "all metals at background level 1 μg/L" vs.
"all metals at EPA action level 5 μg/L". Because `prepare_gbkmr_data()`
log-transforms exposures by default, you must pass values on the
*transformed* scale (`log(1) = 0`, `log(5) ≈ 1.609`):

```{r custom}
# Get exposure column names (the 6 columns vary by visit)
exposure_cols <- paste0("logM", rep(1:2, 3), "_", rep(0:2, each = 2))
# c("logM1_0", "logM2_0", "logM1_1", "logM2_1", "logM1_2", "logM2_2")

a_background <- setNames(rep(log(1), 6), exposure_cols)
a_action     <- setNames(rep(log(5), 6), exposure_cols)

# For custom contrast, use run_gbkmr_panel() (gives full control)
results_regulatory <- run_gbkmr_panel(
  sim_popn = prepared_data,
  T = 3, p = 2,
  mediator_basenames = "waist",
  common_covariates = c("sex", "waist_0"),
  a_vals = a_background,
  astar_vals = a_action,
  iter = 15000, n = 500, K = 1000, n_knots = 50
)
```

You can also mix scenarios: e.g., metal 1 at background but metal 2 at
action level:

```{r mixed}
a_mixed_low  <- setNames(c(log(1), log(1), log(1), log(1), log(1), log(1)), exposure_cols)
a_mixed_high <- setNames(c(log(1), log(5), log(1), log(5), log(1), log(5)), exposure_cols)
#                          ^M1_0   ^M2_0   ^M1_1   ^M2_1   ^M1_2   ^M2_2
# Only M2 increases between scenarios
```

## What NOT to do

A common pitfall is to put **non-mixture variables** into `a_vals` /
`astar_vals`:

```{r wrong}
# WRONG
a_wrong <- c(logM1_0 = -1, logM2_0 = -1, waist_0 = 80)   # <-- waist_0 should not be here!
```

`waist_0` is a baseline confounder. Changing it between scenarios means
you would be estimating the effect of changing the **confounder**, not
the mixture. The Input Audit prints which columns will vary; verify
before launching a long MCMC run.

## Comparing contrasts

For policy questions, it can be helpful to run **multiple contrasts**
and compare:

```{r compare}
# Three contrasts: mild, moderate, strong shifts
contrasts <- list(
  mild     = c(0.25, 0.75),
  moderate = c(0.10, 0.90),
  strong   = c(0.05, 0.95)
)

results_list <- lapply(names(contrasts), function(nm) {
  gbkmr_run(
    data = prepared_data, time_points = 3,
    a_probs = contrasts[[nm]],
    iter = 15000
  )
})

# Compare estimates
do.call(rbind, lapply(seq_along(results_list), function(i) {
  data.frame(
    contrast = names(contrasts)[i],
    ACE = results_list[[i]]$causal_effect$estimate,
    lower = results_list[[i]]$causal_effect$lower,
    upper = results_list[[i]]$causal_effect$upper
  )
}))
```

## Summary

- `a_probs` is the simplest way to change the contrast.
- `a_vals` / `astar_vals` (in `run_gbkmr_panel()`) gives full control.
- Always restrict the contrast to **mixture columns only**.
- Check the Input Audit before launching MCMC.

## See also

- [Methodology overview](gbkmr_method_overview.html)
- [Quick Start](gbkmr_quickstart.html)
