---
title: "Example 3: Large samples with the fastBKMR engine"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gbkmr_example_fastbkmr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Standard BKMR scales as $O(n^3)$ per MCMC iteration. For cohorts of MESA
or ARIC scale ($n = 6{,}000$+), a single g-BKMR run can take 30+ hours.
The **fast-BKMR engine** (Sonabend et al., 2024) accelerates this by:

1. Splitting the data into $M$ random subsets of size $\approx n/M$.
2. Fitting standard BKMR independently on each subset (optionally in
   parallel across cores).
3. Combining posterior summaries via median aggregation (WASP).

In our benchmark (Use Case 3 of the paper), this yields ~15× speed-up
at $n = 6{,}000$ with near-identical posterior beta estimates.

## When to use fastBKMR

| Sample size      | Recommended engine                   |
|------------------|--------------------------------------|
| $n \le 2{,}000$  | `bkmr` (with knot approximation)     |
| $n > 2{,}000$    | `fastbkmr` (auto-selected by default) |

The `engine = "auto"` default uses the above rule.

## Installation

`fbkmr` is not on CRAN, so install from GitHub:

```{r install-fbkmr}
remotes::install_github("junwei-lu/fbkmr")
```

## Running with fastBKMR (automatic)

If your prepared data has $n > 2{,}000$, the auto engine picks fastBKMR.
No code change needed:

```{r auto}
library(causalBKMR)

# Suppose prepared_data has 6000 rows
results_fast <- gbkmr_run(
  data = prepared_data,
  time_points = 3,
  iter = 15000
  # engine = "auto" (default), n_subset and n_cores auto-set
)
```

The Input Audit will print:

```
Auto-selected engine: fastbkmr (n = 6000 > 2000)
...
[Model configuration]
  Engine:          fastbkmr
  fastBKMR setup:  6 subsets x ~1000 obs, 6 cores
```

By default the package uses $M = \lfloor n / 1000 \rfloor$ subsets
(at least 5) and the lesser of $M$, available cores, or 10 for
`n_cores`.

## Forcing fastBKMR at smaller $n$

Even at $n = 1{,}500$ you can force fastBKMR if you have many cores:

```{r force}
results_force <- gbkmr_run(
  data = prepared_data,
  time_points = 3,
  engine = "fastbkmr",
  n_subset = 5,
  n_cores = 5,
  iter = 15000
)
```

## How accurate is fastBKMR?

In our paper benchmark (Use Case 3), $n = 6{,}000$ over 3 fitted models:

| Method          | Beta (sex, waist₀) | Time   |
|-----------------|---------------------|--------|
| BKMR ($n$=500)  | (0.027, 0.492)      | 6 min  |
| BKMR ($n$=6K)   | (-0.005, 0.517)     | 724 min |
| fastBKMR ($n$=6K)| (-0.000, 0.515)    | 47 min |

- Beta estimates are **statistically indistinguishable** between full
  BKMR and fastBKMR at the same $n$.
- Variable selection (PIPs) is correctly recovered, though slightly
  less sharp than full BKMR — an expected consequence of subset
  splitting and WASP aggregation.

## Limitations

- **Binary outcome with fastBKMR is not supported in v1.** The
  `fbkmr::skmbayes()` function does not expose a `family` argument.
  See [Example 2](gbkmr_example_binary.html) for binary outcomes
  (use standard BKMR engine).
- **No knot approximation in fastBKMR.** This is a deliberate choice
  by the `fbkmr` authors (double approximation reduces accuracy).
  As a result, fastBKMR's per-subset cost still scales with the number
  of $Z$ columns; expect ~5–25 min per model at $n = 6{,}000$.
- **Parallelisation requires `doSNOW` + `doParallel`.** On HPC clusters
  these are typically pre-installed; on a personal machine, run
  `install.packages(c("doSNOW", "doParallel"))` before using fastBKMR
  in parallel.
- **Login-node parallelism may fail.** Some HPC login nodes do not
  allow spawning subprocesses. The package automatically falls back
  to sequential mode if `doParallelSNOW` cannot start; see the
  warning printed in such cases.

## Comparing the two engines on your data

For sample sizes near the threshold ($1{,}500 \le n \le 3{,}000$),
it's worth running both engines as a sanity check:

```{r compare}
res_bkmr     <- gbkmr_run(data = prepared, time_points = 3,
                          engine = "bkmr", iter = 15000)
res_fastbkmr <- gbkmr_run(data = prepared, time_points = 3,
                          engine = "fastbkmr", iter = 15000)

# Should be very close
res_bkmr$causal_effect
res_fastbkmr$causal_effect
```

## See also

- [Methodology overview](gbkmr_method_overview.html) — divide-and-conquer description
- [Quick Start](gbkmr_quickstart.html)
- Sonabend, A. et al. (2024). arXiv:2411.10858
